---
title: "Fish Stocked in the Great Lakes"
format:
  html:
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

## Data Description

1.  The source of the data is from the Tidy Tuesday package that has data on fish stocking in the Great Lakes found from the Great Lakes Fishery Commission 1867-2015.
2.  The data includes variables such as the year, month, lake, species, state, and number of fish stocked.
3.  The research questions I would like to address includes Which lakes get stocked the most and do the species of fish differ by each lake? The second question is Has the number of fish stocked changed over the years and is there a seasonal difference?

## Data Visualization

I want my final visualizations to show the trends over time using a line plot and highlight each of the seasons using different lines. I would like another visualization to show the number of fish stocked for each lake using something like a bar chart that highlights each of the species types.

## Data Cleaning

1.  I need to create a season variable by grouping the months into each of the 4 seasons.
2.  I need to remove a couple of variables that I am not interested in.
3.  I would also need to filter out the N/A values that are missing in the months variable for when I am creating seasons.


```{r}
#| label: load-packages
#| message: false
#| echo: false
#| warning: false
#| include: false

library(janitor)
library(tidyverse)
library(plotly)
library(fishpals)
```

```{r}
#| label: read-data
#| message: false
#| warning: false
#| echo: false
#| include: false

stocked <- readr::read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-06-08/stocked.csv",
  show_col_types = FALSE
)

stocked <- janitor::clean_names(stocked)

my_theme <- function() {
  list(
    theme_bw(),
    scale_color_fishpals("adultchinook", discrete = TRUE),
    scale_fill_fishpals("adultchinook", discrete = TRUE)
  )
}
```

```{r}
#| label: clean-data
#| message: false
#| warning: false
#| echo: false
#| include: false

stocked_clean <- stocked |>
  filter(!is.na(month), year > 1980) |>
  mutate(
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% c(3, 4, 5) ~ "Spring",
      month %in% c(6, 7, 8) ~ "Summer",
      month %in% c(9, 10, 11) ~ "Fall"
    )
  ) |>
  select(year, month, lake, no_stocked, species, season) |>
  group_by(lake) |>
  mutate(total_no_stocked = sum(no_stocked, na.rm = TRUE),
         total_stocked_mill = total_no_stocked / 1000000,
         total_stocked_k = total_no_stocked / 1000) |>
  ungroup() 

stocked_clean <- stocked_clean |>
  mutate(species_group = case_when(
    species %in% c("RBT", "LAT", "BNT", "BKT", "SPE") ~ "Trout",
    species %in% c("CHS", "COS", "ATS")                ~ "Salmon",
    species == "WAE"                                    ~ "Walleye",
    species == "MUE"                                    ~ "Muskellunge",
    species == "YEP"                                    ~ "Yellow Perch",
    .default = "Other"
  ))

```

## Visualization 1


```{r}
#| label: fish-stock-lake
#| message: false
#| warning: false
#| echo: false
#| fig-cap: "Michigan and Ontario had the highest total fish stocked from 1980–2015, with trout dominating across most lakes. Across all lakes, the species Salmon, Trout, and Walleye were stocked the most."
#| fig-alt: "A stacked bar chart showing total fish stocked in millions for each of the Great Lakes, ordered from highest to lowest. Lakes are on the x-axis and total fish stocked in millions is on the y-axis. Bars are filled by species group including trout, salmon, walleye, muskellunge, yellow perch, and other. Michigan and Ontario have the tallest bars."


stocked_clean |>
  summarise(total = sum(no_stocked), .by = c(lake, species_group)) |>
  ggplot(aes(x = reorder(lake, -total, sum), y = total / 1e6, fill = species_group)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Total Fish Stocked by Lake and Species Group",
    x = "Lake",
    y = "Total Fish Stocked (millions)",
    fill = "Species Group"
  ) +
  my_theme()
```

## Visualization 2


```{r}
#| label: stock-years
#| message: false
#| warning: false
#| echo: false
#| fig-cap: "Fish stocking in the Great Lakes peaked around the mid-1980s and declined through 2015. Spring stocking consistently outpaced summer, fall, and winter across the entire period."
#| fig-alt: "A line chart showing total fish stocked in millions per year from 1980 to 2015, with separate colored lines for each of the four seasons. Spring and summer lines are generally higher than fall and winter. All four lines show a peak in the mid-1980s followed by a gradual decline."

stocked_clean |>
  summarise(total = sum(no_stocked), .by = c(year, season)) |>
  ggplot(aes(x = year, y = total / 1e6, color = season)) +
  geom_line() +
  labs(
    title = "Fish Stocked Over the Years by Season",
    x = "Year",
    y = "Total Fish Stocked (millions)",
    color = "Season"
  ) +
  my_theme()
```












